---
title: "How ‘Love’ Appears in Romeo & Juliet"

format: html

execute:

  echo: true

  warning: false

  message: false
---

# Introduction 

This analysis examines the frequency and distribution of the word *“love”* in the dialogue of *Romeo and Juliet*, identifying which characters use it most often and in what contexts. Its goal is to quantify the term’s explicit usage and trace patterns in how it appears across the play. The project also highlights key moments where *“love”* is paired with *“death,”* showing how Shakespeare uses this juxtaposition to reinforce the play’s tragic tone. Specifically, it addresses the following questions:

-   How many lines explicitly contain the word “love”?

-   Which characters use the word most frequently?

-   In which lines does Shakespeare pair “love” with “death,” and what might these instances suggest about the play’s tragic tone?

# Data & Preprocessing 

The data come from the TidyTuesday 2024‑09‑17 release, which provides a tidy version of *Romeo & Juliet* with one row per line of dialogue and columns for the scene, speaking character, and text of the line.

Several new variables are created to encode whether and how “love” appears in each line:

-   `has_love`: line contains the word “love” (case‑insensitive, whole word).

-   `love_count`: number of times “love” appears in the line.

-   `starts_love`: line begins with the word “love."

-   `love_and_death`: line contains “love” and, later in the same line, “death.”

These indicators let the analysis move from raw text to counts and visual summaries.

The stringr functions used in the analysis include:

-   `str_detect()` : creates logical indicators for specific patterns -

    -   `str_detect(dialogue, "(?i)\\blove\\b")` checks whether a line contains the whole word “love” and is used to define `has_love`.

    -   `str_detect(dialogue, "^(?i)love")` identifies lines that begin with “love” and defines `starts_love`.

    -   `str_detect(dialogue, "(?i)love(?=.*death)")` detects lines where “love” appears and “death” occurs later in the same line, giving `love_and_death`.

-   `str_count()` : counts pattern occurrences -

    -   `str_count(dialogue, "(?i)\\blove\\b")` returns the number of times “love” appears in each line, stored in `love_count`.

-   `str_replace_all()` is used to highlight “love” in examples -

    -   `str_replace_all(dialogue, "(?i)\\blove\\b", "LOVE")` constructs `dialogue_highlight`, a version of each line where “love” is capitalized to stand out.

```{r}
library(tidyverse) 
library(stringr) 
library(tidytuesdayR)

# Download the Romeo & Juliet dataset 
tuesdata <- tt_load(2024, week = 38) 
# Extract the romeo_juliet data frame from the list 
romeo_juliet <- tuesdata$romeo_juliet
```

## How Often is “Love” Used?

#### Figure 1 : Lines Mentioning 'Love' in Romeo & Juliet

```{r}
# Add indicators and counts for the word "love"
romeo_juliet <- romeo_juliet %>%
  mutate(
    has_love   = str_detect(dialogue, "(?i)\\blove\\b"),
    love_count = str_count(dialogue, "(?i)\\blove\\b")
  )

romeo_juliet %>%
  count(has_love) %>%                         # count lines by whether they contain "love"
  ggplot(aes(x = has_love, y = n, fill = has_love)) +
  geom_col(show.legend = FALSE) +
  scale_x_discrete(labels = c("No", "Yes")) +
  labs(
    title = "Lines Mentioning 'Love' in Romeo & Juliet",
    x = "Line contains the word 'love'?",
    y = "Number of lines"
  ) +
  theme_minimal()
```

This bar plot compares the number of lines that do and do not contain the word “love” in *Romeo & Juliet*. Only a modest fraction of lines explicitly use the word, indicating that although love is a central theme, much of it is conveyed indirectly through actions, imagery, and other language rather than repeated direct mentions of “love.”

## Which Characters Speak of Love Most?

To identify who explicitly names “love,” the data are restricted to lines where `has_love` is TRUE, and counts are computed by character.

#### Figure 2 : Characters with the most “love” lines

```{r}
top_love <- romeo_juliet %>%
  filter(has_love, !is.na(character)) %>%  # keep only "love" lines with a known speaker
  group_by(character) %>%
  tally(sort = TRUE)      

top_love %>%
  slice_head(n = 8) %>%                    # show top 8 characters
  ggplot(aes(x = reorder(character, n), y = n, fill = character)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  labs(
    title = "Characters with the Most 'Love' Lines",
    x = "Character",
    y = "Number of lines containing 'love'"
  ) +
  theme_minimal()
```

This horizontal bar chart shows the characters with the largest number of lines that include the word “love.” While the total number of such lines is limited, this plot highlights which speakers most often name “love” explicitly, providing a simple view of who voices the theme in dialogue and how strongly it is tied to particular characters.

## Contexts of “Love”

The next step focuses on more specific textual contexts: lines that start with “love,” and lines that explicitly mention both “love” and “death.”

#### Table 1: Example lines where both "love" and "death" appear

```{r}
romeo_juliet <- romeo_juliet %>%
  mutate(
    starts_love = str_detect(dialogue, "^(?i)love"),
    love_and_death = str_detect(dialogue, "(?i)love(?=.*death)")
  )


#  Lines where both "love" and "death" appear

love_death_lines <- romeo_juliet %>%
  # Keep only lines flagged as having both "love" and "death"
  filter(love_and_death) %>%
  # Show just the speaker and the original dialogue text
  select(character, dialogue)

love_death_lines

```

This table lists the lines in which “love” and “death” occur in the same sentence, with “love” appearing before “death.” There are only a few such lines, but they vividly illustrate moments where Shakespeare explicitly links passion and mortality. The sparing use of this combination helps these lines stand out and supports the tragic atmosphere and foreshadowing that run through the play.

## Sentiment in Love Lines

To visualize “love” within its textual context, the analysis highlights each occurrence of the word in all caps.

#### **Table 2: Sample Lines Highlighting “LOVE” in Context**

```{r}
romeo_juliet_sample <- romeo_juliet %>%
  filter(has_love) %>%   # keep lines that contain "love"
  mutate(
    dialogue_highlight = str_replace_all(
      dialogue,
      "(?i)\\blove\\b",
      "LOVE"
    )
  )

romeo_juliet_sample %>%
  select(character, dialogue_highlight) %>%
  head(5)
```

This sample table shows several lines that contain “LOVE,” with the word capitalized to draw attention to its position in each sentence. The highlighting does not alter the wording but makes it easier to see how “love” is embedded in different emotional contexts. These examples complement the earlier counts and plots by showing the actual
language that underlies the summaries.

## References/Data Source

The analysis uses the TidyTuesday “Romeo & Juliet” text dataset, which provides one row per line of dialogue with scene and character information.

**TidyTuesday dataset**

Rfordatascience. (2024, September 17). *tidytuesday/data/2024/2024-09-17/romeo_juliet.csv at main · rfordatascience/tidytuesday*. GitHub. https://github.com/rfordatascience/tidytuesday/blob/main/data/2024/2024-09-17/romeo_juliet.csv

**Project Gutenberg**

*Project Gutenberg*. (n.d.). Project Gutenberg. <https://www.gutenberg.org/>
